---
title: "Label retrieval"
description: "Choose ROC or precision-recall AUC and inspect overall, per-label, and per-query retrieval scores."
---

Use this guide when known labels define what should be retrieved near each
observation. Choose ROC AUC for rank discrimination across relevant and
irrelevant candidates. Prefer PR AUC when relevant observations are rare and
precision among retrieved observations matters.

`roc_auc()` and `pr_auc()` treat each observation as a query, rank the remaining
observations by distance, and regard matching labels as relevant. These are
supervised retrieval summaries. Both functions require the
[PRROC package](https://cran.r-project.org/package=PRROC).

Perfect retrieval has AUC 1 for both measures. Random ROC AUC is 0.5. The random
PR AUC baseline is the prevalence of relevant observations among the remaining
candidates, so it changes with label imbalance.

```{r label-retrieval-setup}
library(quadra)

iris_x <- as.matrix(iris[, -5])
pca_iris <- stats::prcomp(iris_x, retx = TRUE, rank. = 2)$x
dout <- as.matrix(stats::dist(pca_iris))
```

```{r overall-and-label-auc}
roc <- roc_auc(dout, iris$Species)
pr <- pr_auc(dout, iris$Species)

auc_summary <- rbind(
  ROC = c(overall = roc$av_auc, unlist(roc$label_av)),
  PR = c(overall = pr$av_auc, unlist(pr$label_av))
)
signif(auc_summary, 5)
```

```{r overall-and-label-auc-check, include=FALSE}
iris_pr_baseline <- (table(iris$Species) - 1) / (nrow(iris) - 1)
stopifnot(
  identical(colnames(auc_summary), c("overall", levels(iris$Species))),
  all(is.finite(auc_summary)),
  all(auc_summary >= 0 & auc_summary <= 1),
  roc$av_auc > 0.5,
  all(unlist(pr$label_av) > iris_pr_baseline),
  all(auc_summary[, "setosa"] > 0.999),
  all(auc_summary[, "setosa"] < 1),
  all(
    sweep(
      auc_summary[, c("versicolor", "virginica"), drop = FALSE],
      1,
      auc_summary[, "setosa"],
      FUN = "<"
    )
  )
)
```

Every Iris class has 50 observations, so each query has 49 relevant candidates
among the other 149 and a random PR baseline of about 0.329. The overall Iris
scores are above their random baselines. Setosa is retrieved almost perfectly
in this PCA embedding; for versicolor and virginica, same-label observations
are less consistently ranked ahead of the other classes. The difference
between ROC and PR values is not itself a winner-takes-all comparison: the
measures use different baselines and emphasize different retrieval behavior.

## Inspecting Query-Level Results

By default, each function returns `list(av_auc, label_av)`. `av_auc` weights
each defined query equally; `label_av` contains per-label means in
first-appearance order. Set `ret_extra = TRUE` to append the row-aligned
`query_auc` vector and its defined-query count.

```{r query-level-auc}
roc_detail <- roc_auc(dout, iris$Species, ret_extra = TRUE)
head(round(roc_detail$query_auc, 3))
roc_detail$n_defined
```

```{r query-level-auc-check, include=FALSE}
stopifnot(
  length(roc_detail$query_auc) == nrow(iris),
  roc_detail$n_defined == nrow(iris),
  all(is.finite(roc_detail$query_auc)),
  all(roc_detail$query_auc >= 0 & roc_detail$query_auc <= 1)
)
```

All 150 Iris queries are defined. In other data, queries without both relevant
and irrelevant candidates are `NA_real_` and excluded from the overall and
per-label means. If no query is defined, the overall average is `NA_real_`.

Both functions require a finite square distance matrix and one nonmissing label
per observation. The query itself is excluded from its ranking. Labels without
a defined query remain `NA_real_` in `label_av`.

## Further Reading

Venna, J., Peltonen, J., Nybo, K., Aidos, H., & Kaski, S. (2010).
Information retrieval perspective to nonlinear dimensionality reduction for data visualization.
*Journal of Machine Learning Research*, *11*(Feb), 451-490.
<http://www.jmlr.org/papers/v11/venna10a.html>

### Precision-Recall AUC and Receiver Operating Characteristic Area Under the Curve

Davis, J., & Goadrich, M. (2006, June).
The relationship between Precision-Recall and ROC curves.
In *Proceedings of the 23rd international conference on Machine learning*
(pp. 233-240). ACM.
<http://pages.cs.wisc.edu/~jdavis/davisgoadrichcamera2.pdf>

Keilwagen, J., Grosse, I., & Grau, J. (2014).
Area under precision-recall curves for weighted and unweighted data.
*PloS One*, *9*(3), e92209.
<https://dx.doi.org/10.1371/journal.pone.0092209>
