---
title: "Local preservation"
description: "Choose and interpret metrics for neighbor identity, rank, local scale, graph structure, and multiple neighborhood sizes."
---

Use this guide to choose a local metric from the relationship that matters for
your analysis. Start with `nn_preservation()` when neighbor identity is the
main concern. Use trustworthiness or continuity for rank-sensitive penalties,
local radius correlation for relative neighborhood scale, mutual-neighbor
correlation for reciprocal graph structure, and RNX to inspect agreement over
multiple neighborhood sizes.

The examples use exact brute-force neighbors so their results are deterministic
on this small dataset.

```{r local-setup}
library(quadra)

iris_x <- as.matrix(iris[, -5])
pca_iris <- stats::prcomp(iris_x, retx = TRUE, rank. = 2)$x
```

## Neighborhood Preservation

Nearest-neighbor preservation is a useful default: it reports the mean fraction
of input neighbors retained in the embedding. Pass a vector to `k` to inspect
more than one neighborhood size.

```{r nearest-neighbor-preservation}
neighbor_k <- c(15, 30)
neighbor_scores <- nn_preservation(
  iris_x,
  pca_iris,
  k = neighbor_k,
  nn_method_in = "brute",
  nn_method_out = "brute",
  n_threads = 1
)
chance_baselines <- neighbor_k / (nrow(iris_x) - 1)
round(
  data.frame(
    k = neighbor_k,
    preservation = unname(neighbor_scores),
    chance_baseline = chance_baselines
  ),
  3
)
```

```{r nearest-neighbor-preservation-check, include=FALSE}
stopifnot(
  all(is.finite(neighbor_scores)),
  all(neighbor_scores >= 0 & neighbor_scores <= 1),
  all(neighbor_scores > chance_baselines)
)
```

For this two-dimensional Iris PCA, both preservation values are substantially
above their scale-specific chance baselines. The raw values answer different
neighborhood-scale questions, so the larger value at `k = 30` should not be
read as evidence that 30 neighbors is universally preferable to 15.

## Trustworthiness and Continuity

`trustworthiness()` and `continuity()` are exact rank-penalty neighborhood
metrics for distance matrices. Trustworthiness penalizes observations that
appear as low-dimensional neighbors even though they are not close in the input
space. Continuity applies the dual penalty to input-space neighbors that are no
longer nearby in the embedding.

```{r trustworthiness-continuity}
din <- as.matrix(stats::dist(iris_x))
dout <- as.matrix(stats::dist(pca_iris))
rank_k <- c(5, 15, 30)
trust_scores <- trustworthiness(din, dout, k = rank_k)
continuity_scores <- continuity(din, dout, k = rank_k)
round(
  data.frame(
    k = rank_k,
    trustworthiness = unname(trust_scores),
    continuity = unname(continuity_scores)
  ),
  3
)
```

```{r trustworthiness-continuity-check, include=FALSE}
stopifnot(
  all(c(trust_scores, continuity_scores) >= 0),
  all(c(trust_scores, continuity_scores) <= 1),
  min(c(trust_scores, continuity_scores)) > 0.9
)
```

Values nearer 1 mean smaller rank penalties. Both metrics are above 0.9 at the
shown scales, so this embedding introduces relatively small trustworthiness and
continuity penalties for these Iris neighborhoods.

These functions use full `n` by `n` distance matrices and require `k` to be
less than half the number of observations, so they are intended for exact
small-dataset checks. For larger datasets, use nearest-neighbor preservation
metrics such as `nn_preservation()` or `nbr_pres_knn()`.

## Local Radius Correlation

Nearest neighbor preservation asks whether the same observations remain nearby.
`local_radius_correlation()` asks a different local-scale question: do points in
dense or sparse regions of the input data still have relatively small or large
nearest-neighbor radii in the embedding?

```{r local-radius-correlation}
radius_scores <- local_radius_correlation(
  iris_x,
  pca_iris,
  k = c(15, 30),
  nn_method_in = "brute",
  nn_method_out = "brute",
  n_threads = 1
)
round(radius_scores, 3)
```

```{r local-radius-correlation-check, include=FALSE}
stopifnot(
  identical(names(radius_scores), c("lrc15", "lrc30")),
  all(is.finite(radius_scores)),
  all(radius_scores >= -1 & radius_scores <= 1),
  all(radius_scores > 0.8)
)
```

The strong positive correlations show that observations with relatively small
or large input-space radii tend to retain that relative scale in this Iris
embedding. This conclusion concerns scale, not which observations are
neighbors.

By default, the metric uses Spearman correlation between the distance to the
`k`th nearest non-self neighbor in the input data and output embedding. Set
`statistic = "mean"` to use the mean distance to the first `k` neighbors, or
`log = TRUE` to compare log radii when all selected radii are positive.

With exact neighbors, squared Euclidean preserves the default Spearman ordering
of `k`th-neighbor radii. It can still change Pearson correlations,
mean-neighbor summaries, or approximate-search results.

If you already have nearest-neighbor graphs with distance matrices, you can
reuse them:

```{r reuse-local-radius-graphs}
cached <- local_radius_correlation(
  iris_x,
  pca_iris,
  k = c(15, 30),
  nn_method_in = "brute",
  nn_method_out = "brute",
  n_threads = 1,
  ret_extra = TRUE
)

cached_scores <- local_radius_correlation(
  cached$nn_in,
  cached$nn_out,
  k = c(15, 30),
  n_threads = 1
)
round(cached_scores, 3)
```

```{r reuse-local-radius-graphs-check, include=FALSE}
stopifnot(isTRUE(all.equal(cached$lrc, cached_scores)))
```

The reused graphs give the same result because they retain both the indices and
distances required by the radius calculation.

This metric is a radius or scale diagnostic. It does not estimate local density
directly and should be interpreted alongside neighbor-identity metrics such as
`nn_preservation()`.

## Mutual Neighbor Correlation

`mutual_neighbor_correlation()` compares mutual-neighbor count patterns. For
each observation, the mutual-neighbor count is the number of its first `k`
nearest neighbors that also include the observation among their first `k`
nearest neighbors.

```{r mutual-neighbor-correlation}
mutual_scores <- mutual_neighbor_correlation(
  iris_x,
  pca_iris,
  k = c(15, 30),
  nn_method_in = "brute",
  nn_method_out = "brute",
  n_threads = 1
)
round(mutual_scores, 3)
```

```{r mutual-neighbor-correlation-check, include=FALSE}
stopifnot(
  identical(names(mutual_scores), c("mnc15", "mnc30")),
  all(is.finite(mutual_scores)),
  all(mutual_scores >= -1 & mutual_scores <= 1),
  all(mutual_scores > 0)
)
```

Both correlations are positive, so observations with high or low reciprocal
neighbor counts in the input tend to retain that pattern in this embedding.
The result is about graph structure, not direct neighbor overlap.

This is a graph diagnostic. It does not require neighbor distances, so cached
idx-only nearest-neighbor graphs are enough. It should be interpreted alongside
neighbor-identity metrics such as `nn_preservation()` and scale metrics such as
`local_radius_correlation()`.

## Multi-Scale RNX Agreement

`rnx_auc()` summarizes exact rank-based neighborhood agreement across all
available neighborhood sizes, weighting smaller neighborhoods more heavily.
Use `rnx_curve()` when the scale-specific pattern matters.

```{r rnx-agreement}
rnx_scores <- c(
  auc = rnx_auc(din, dout),
  rnx_curve(din, dout, k = c(5, 15, 30))
)
round(rnx_scores, 3)
```

```{r rnx-agreement-check, include=FALSE}
stopifnot(
  identical(names(rnx_scores), c("auc", "rnx5", "rnx15", "rnx30")),
  all(is.finite(rnx_scores)),
  all(rnx_scores <= 1),
  rnx_scores[["auc"]] > 0,
  all(diff(rnx_scores[c("rnx5", "rnx15", "rnx30")]) > 0)
)
```

An RNX value of 1 is perfect, 0 is the random-neighborhood baseline, and
negative values indicate worse-than-random agreement. The Iris curve rises over
the displayed values of `k`, so this embedding preserves rank neighborhoods
better at the broader shown scales. That pattern diagnoses this embedding; it
does not establish a universally preferred embedding or neighborhood size.

## Using Existing Neighbor Graphs

For larger datasets, `nn_preservation()` can use approximate neighbors through
[rnndescent](https://github.com/jlmelville/rnndescent). If another tool already
provides one-based nearest-neighbor index matrices, pass them directly to
`nbr_pres_knn()`. Rows are observations and columns are neighbors in
nearest-first order.

```{r supplied-neighbor-indices}
input_idx <- matrix(
  c(2, 3, 1, 3, 2, 1, 5, 3, 4, 3),
  ncol = 2,
  byrow = TRUE
)
output_idx <- matrix(
  c(2, 4, 1, 3, 2, 5, 5, 1, 4, 3),
  ncol = 2,
  byrow = TRUE
)

external_overlap <- nbr_pres_knn(
  input_idx,
  output_idx,
  k = 2,
  n_threads = 1
)
external_overlap
```

```{r supplied-neighbor-indices-check, include=FALSE}
stopifnot(
  length(external_overlap) == nrow(input_idx),
  all(external_overlap >= 0 & external_overlap <= 1),
  sum(external_overlap == 1) == 2,
  sum(external_overlap == 0.5) == 3
)
```

Two rows retain both neighbors and the other three retain one of two. This
shows how to reuse one-based neighbor indices from any graph-construction tool
without adding an adapter dependency.

Supplied graphs use one-based indices in nearest-first order. Quadra checks all
columns, removes self-indices, and retains the first `k` non-self neighbors, so
self-inclusive graphs need at least `k + 1` columns. Supplied order resolves
ties; `nbr_pres()` includes boundary ties, while exact rank metrics break ties
by distance-matrix column order.

Graph searches pass supported sparse inputs to `rnndescent`; nonnumeric
data-frame columns are ignored. Exact Euclidean and squared-Euclidean searches
have the same ordering, subject to ties and approximation.

## Further Reading

### Neighbor Overlap and Rank Criteria

France, S., & Carroll, D. (2007, July).
Development of an agreement metric based upon the RAND index for the evaluation
of dimensionality reduction techniques, with applications to mapping customer
data.
In *International Workshop on Machine Learning and Data Mining in Pattern Recognition*
(pp. 499-517). Springer, Berlin, Heidelberg.
<https://doi.org/10.1007/978-3-540-73499-4_38>

Chen, L., & Buja, A. (2009). Local multidimensional scaling for nonlinear
dimension reduction, graph drawing, and proximity analysis.
*Journal of the American Statistical Association*, *104*(485), 209-219.
<http://dx.doi.org/10.1198/jasa.2009.0111>

Lee, J. A., & Verleysen, M. (2009).
Quality assessment of dimensionality reduction: Rank-based criteria.
*Neurocomputing*, *72*(7), 1431-1443.
<https://dx.doi.org/10.1016/j.neucom.2008.12.017>

Lee, J. A., Peluffo-Ordonez, D. H., & Verleysen, M. (2015).
Multi-scale similarities in stochastic neighbour embedding: Reducing
dimensionality while preserving both local and global structure.
*Neurocomputing*, *169*, 246-261.
<https://dx.doi.org/10.1016/j.neucom.2014.12.095>

### Local Scale and Density Context

Cooley, S. M., Hamilton, T., Deeds, E. J., & Ray, J. C. J. (2019).
A novel metric reveals previously unrecognized distortion in dimensionality reduction of scRNA-Seq data.
*BioRxiv*, 689851.
<https://www.biorxiv.org/content/10.1101/689851v6>

Narayan, A., Berger, B., & Cho, H. (2021).
Assessing single-cell transcriptomic variability through density-preserving
data visualization.
*Nature Biotechnology*, *39*, 765-774.
<https://doi.org/10.1038/s41587-020-00801-7>
